Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2020-07-25 paper2_JNLS/paper.tex
NM

%----------------------------------------------------------------------------%
%% LATEX DEFINITIONS
%----------------------------------------------------------------------------%
% Basic editing
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
\newcommand{\nick}[1]{{\color{red} #1}}
\newcommand{\new}[1]{{\color{blue} #1}}
% Derivatives
\newcommand{\pd}[2]    { \frac{\partial #1} {\partial #2} }
\newcommand{\ppd}[2]  { \frac{\partial^2 #1}{{\partial #2}^2} }
\newcommand{\pdi}[2] { {\partial_#2} #1 }
\newcommand{\td}[2] { \frac{d #1} { d #2 } }
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
% Vectors and operators
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\eps}{\varepsilon}
\newcommand{\dx}{\, dx}
\newcommand{\CC}{{\mathbb{C}}}

@@

% ORIGINAL: New experimental measurements confirm this prediction in spectacular fashion.

% OTHER
% New experimental measurements provide convincing evidence for this prediction.
% convincing, conclusive, compelling.

@@

In this paper, we only examine statistic of mean-zero quantities, and so, for an arbitrary mean-zero quantity $q$, we have the following definitions
\begin{align}
%& \var(q) = \mean{q^2}		&&\mbox{\em variance} \\
& \std(q) = q_{std}= \sqrt{\mean{q^2}}
&&\mbox{\em standard deviation} \\
& \skw(q) = {\mean{q^3}} / {q_{std}^3}	
&&\mbox{\em skewness} \\
& \kurt(q) = {\mean{q^4}} / {q_{std}^4} - 3
&&\mbox{\em (excess) kurtosis}
\end{align}
where $\mean{}$ indicates a mean --- here a long-time mean at a fixed spatial location. Hereafter, we will simply refer to the excess kurtosis as kurtosis.

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l l}
%\hline \multicolumn{3} { c }{Parameters that are constant in a single experiment} \\
\hline Description & Notation & Experimental range & `Representative' values \\
\hline
Peak forcing frequency	& $f_p$					& 2 Hz 		& 2 Hz \\
Characteristic amplitude	& $\etastd$				& 0.03--0.3 cm	& 0.21 cm \\
Upstream depth		& $\dup$					& 12.5 cm 	& 12.5 cm \\
Downstream depth		& $\ddn$					& 2.2--5.3 cm	& 3 cm \\
Upstream wavelength	& $\lamup = \sqrt{g \dup}/f_p$	& 55 cm 		& 55 cm \\
Downstream wavelength	& $\lamdn = \sqrt{g \ddn}/f_p$	& 23--36 cm	& 27 cm \\
%
Amplitude-to-depth ratio	& $\epsup = \etastd / \dup$	&0.0024--0.024	& 0.017 \\
Depth-to-wavelength ratio	& $\delup = \dup / \lamup$	& 0.23		& 0.23 \\
Depth ratio			& $\dratdn = \ddn/\dup$		& 0.18--0.42	& 0.24
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

In the experiments, the randomized incoming wave-field is generated with a peak forcing frequency of $\freqp = 2$ Hz, which gives rise to the characteristic wavelength of $\lam = c/\freqp = \sqrt{g \depth} / \freqp$. Note that both the characteristic wave speed $c = c_{\pm}$ and wavelength $\lam = \lamupdn$ take different values upstream and downstream of the ADC. We remark that experimental measurements indicate that $\etastd$ is nearly the same on both sides of the ADC. Hence, we will not distinguish between upstream and downstream values of $\etastd$.
%See Table \ref{paramtable} for a summary of important parameters.

@@

% UNUSED POISSON BRACKET STUFF
\begin{comment}
The evolution of any functional $\Fcnl[u]$ is given by 
\begin{align}
\partial_t \Fcnl  = \{ \Fcnl, \Ham \} = \int_{-\pi}^{\pi} \vard{\Fcnl}{u} \sympJ \vard{\Ham}{u} \dx
\end{align}
where again $\Ham = \Hup$ if $t<0$, and $\Ham = \Hdn$ if $t>0$ and $\{\}$ is the Poisson bracket induced by $\sympJ$.
\end{comment}

@@

We now introduce the truncated KdV (TKdV) system, which is the main focus of the present study. Consider the state variable represented as a spatial Fourier series
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k(t) \, e^{i k x} \, , \\
\label{uhat}
&\uhat_k(t) = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x,t) \, e^{-i k x} \dx \, ,
\end{align}
where $\uhat_k(t) \in \mathbb{C}$. Since $u(x,t)$ is real valued, $\uhat_{-k} = \uhat_{k}^*$, and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k(t) \, e^{i k x} \, , \qquad
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Inserting the projected variable, $\uL$, into the KdV equation and applying the projection operator, $\Proj$, again where necessary produces the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad C_2 = \frac{2 \pi^2 \delup}{3 \lamfac^2}
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. Since all wavenumbers larger than $\Lambda$ have been removed, \eqref{TKdV} represents a {\em finite} dimensional dynamical system, of dimension $\Lambda$ over $\mathbb{C}$. The constants $C_3$ and $C_2$ are the same as before and have been repeated here for convenience.

@@

	More specifically, we assume continuity of the surface displacement, $\eta$, across the ADC \cite{whitham2011linear, rey1992propagation}. Since the propagation speed $c$ changes with depth, waves crossing the ADC must rapidly adjust in wavelength in order to match the oscillation frequency just upstream of the ADC \cite{whitham2011linear, rey1992propagation}. The normalized domain $x \in [-\pi,\pi]$ is scaled on the characteristic wavelength and so there is no change in the dimensionless wave-field $u(x,t)$, giving the condition
\begin{equation}
\label{detmatch}
u(x,t) \vert_{t=0^-} = u(x,t) \vert_{t=0^+}, 
%\quad \mbox{for } -\pi \le x \le \pi,
\qquad \mbox{\em Deterministic matching condition}
\end{equation}	
We call \eqref{detmatch} the deterministic matching condition to contrast with its statistical counterpart introduced below. This matching condition is employed in the deterministic simulations of TKdV \eqref{TKdV}. We note that in the alternate formulation mentioned earlier, it is the product $\depth^{1/4} \eta$ that would match at the ADC \cite{johnson1997modern}.

@@

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS[\uL]
%\qquad \text{for } x \in [-\pi,\pi]
\end{align}	
where we have represented the right-hand side by the operator $\RHS[\uL]$. We employ a pseudo-spectral discretization of \eqref{TKdV2}, with de-aliasing applied to the quadratic nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3\Lambda/2$, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$.
For simplicity, we denote these Fourier coefficients $\usqdah_k$,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}
With this discretization, \eqref{TKdV2} can be recast in spectral space as a nonlinear ODE system
\begin{equation}
\label{TKdVDisc}
\td{}{t} \uhat_k =  -\frac{1}{2} C_3 \drat^{-3/2} \, ik \usqdah_k + C_2 \drat^{1/2} \, ik^{3} \uhat_k = \RHSh_k
\end{equation}
The quadratic nonlinearity represented by $\usqdah_k$ mixes the modes during evolution. We note the third-order linear term may become stiff for large $\Lambda$.

@@

With the experimental setup described and the theory outlined, we now present results comparing the two. Unless stated otherwise, all parameters used in the theory are taken directly from their experimental values listed in Table \ref{paramtable}. 
%In particular, we set $\delup = 0.23$ in all tests, and for the `representative' experiments, we set $\dratdn = 0.24$ and $\epsup = 0.017$.

@@

	Next, we aim to make the same comparison for free-surface slopes. We note that the derivative, $\partial \eta / \partial x$, is already a dimensionless quantity with a simple physical interpretation, namely the slope of the free surface, whereas the interpretation of $\partial u/\partial \tilde{x}$ is tied to the characteristic values $\ampscale$ and $\lengthscale$.  
We therefore convert the theoretical calculated slopes values back to physical slopes via
\begin{equation}
\pd{\eta}{x} = 2 \pi^{3/2} \frac{\etastd}{\lamupdn} \pd{\uupdn}{\tilde{x}}
%\frac{\ampscale \lamfac}{\lengthscale} \pd{\uupdn}{\tilde{x}}
\end{equation}
So that the peak experimental wavelength $\lam$ (corresponding to $\freqp = 2$ Hz) would be mapped to the dominating TKdV spectral mode, $k = 1$,  this conversion formula has not been corrected with the extra factor of $\lamfac$ present in \eqref{scales}.

@@

%Figure \ref{ExpSpatialStats} may offer additional insight on these detailed differences in the shape of the downstream slope distributions. In particular, \ref{ExpSpatialStats}(e)--(f) show the spatial variation of the slope skewness and kurtosis. Note that variations in these statistics are significantly more transient than the corresponding variations in surface displacement; i.e.~the peaks in surface slope statistics arise more rapidly and decay more rapidly. Variations in displacement statistics also decay, but over a longer spatial scale, presumably due to dissipative effects. The TKdV histograms shown in Fig.~\ref{slopehist}, meanwhile, are long-time averages and hence not intended to capture such transient effects. These observations offer the suggestion that slope dynamics experience stronger transients than do surface displacements, leading to greater difficulty in the direct comparison between measurements extracted from a single location (experiments) and long-time averages (theory).

@@

Inserting \eqref{H3convert} and \eqref{H2convert} into \eqref{H3H2ratio}, and using the definitions of $C_2$ and $C_3$ from \eqref{C3C2}, gives the explicit formula
\begin{equation}
\label{SkewPred}
\frac{\skwdn(\eta)} {\vardn(\eta_x) - \varup(\eta_x)} = \frac{1}{3} \epsup^{-3} \dratdn^3 
%= \frac{1}{3} \left( \frac{\ddn}{\etastd} \right)^3
\end{equation}
This formula links the skewness of the outgoing wave field to the change in the variance of surface slope. In particular, it predicts their ratio to scale as the {\em inverse cube of wave amplitude}, $\epsup^{-3}$, and the {\em cube of the depth ratio}, $\dratdn^3$. It is a rather unexpected relationship, as it would have been difficult to anticipate that the displacement skewness and slope variance, among all possible variable combinations, are so intimately related.

>>>>

%----------------------------------------------------------------------------%
%% LATEX DEFINITIONS
%----------------------------------------------------------------------------%
% Basic editing
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
\newcommand{\nick}[1]{{\color{red} #1}}
\newcommand{\new}[1]{{ #1}}
% Derivatives
\newcommand{\pd}[2]    { \frac{\partial #1} {\partial #2} }
\newcommand{\ppd}[2]  { \frac{\partial^2 #1}{{\partial #2}^2} }
\newcommand{\pdi}[2] { {\partial_#2} #1 }
\newcommand{\td}[2] { \frac{d #1} { d #2 } }
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
% Vectors and operators
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\eps}{\varepsilon}
\newcommand{\dx}{\, dx}
\newcommand{\CC}{{\mathbb{C}}}

@@

@@

In this paper, we only examine statistic of mean-zero quantities, and so, for an arbitrary mean-zero quantity $q$, we have the following definitions
\begin{align}
& \std(q) = q_{std}= \sqrt{\mean{q^2}}
&&\mbox{\em standard deviation} \\
& \skw(q) = {\mean{q^3}} / {q_{std}^3}	
&&\mbox{\em skewness} \\
& \kurt(q) = {\mean{q^4}} / {q_{std}^4} - 3
&&\mbox{\em (excess) kurtosis}
\end{align}
where $\mean{}$ indicates a mean --- here a long-time mean at a fixed spatial location. Hereafter, we will simply refer to the excess kurtosis as kurtosis.

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l l}
\hline Description & Notation & Experimental range & `Representative' values \\
\hline
Peak forcing frequency	& $f_p$					& 2 Hz 		& 2 Hz \\
Characteristic amplitude	& $\etastd$				& 0.03--0.3 cm	& 0.21 cm \\
Upstream depth		& $\dup$					& 12.5 cm 	& 12.5 cm \\
Downstream depth		& $\ddn$					& 2.2--5.3 cm	& 3 cm \\
Upstream wavelength	& $\lamup = \sqrt{g \dup}/f_p$	& 55 cm 		& 55 cm \\
Downstream wavelength	& $\lamdn = \sqrt{g \ddn}/f_p$	& 23--36 cm	& 27 cm \\
%
Amplitude-to-depth ratio	& $\epsup = \etastd / \dup$	&0.0024--0.024	& 0.017 \\
Depth-to-wavelength ratio	& $\delup = \dup / \lamup$	& 0.23		& 0.23 \\
Depth ratio			& $\dratdn = \ddn/\dup$		& 0.18--0.42	& 0.24
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

In the experiments, the randomized incoming wave-field is generated with a peak forcing frequency of $\freqp = 2$ Hz, which gives rise to the characteristic wavelength of $\lam = c/\freqp = \sqrt{g \depth} / \freqp$. Note that both the characteristic wave speed $c = c_{\pm}$ and wavelength $\lam = \lamupdn$ take different values upstream and downstream of the ADC. We remark that experimental measurements indicate that $\etastd$ is nearly the same on both sides of the ADC. Hence, we will not distinguish between upstream and downstream values of $\etastd$.

@@

@@

We now introduce the truncated KdV (TKdV) system, which is the main focus of the present study. Consider the state variable represented as a spatial Fourier series
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k(t) \, e^{i k x} \, , \\
\label{uhat}
&\uhat_k(t) = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x,t) \, e^{-i k x} \dx \, ,
\end{align}
where $\uhat_k(t) \in \mathbb{C}$. Since $u(x,t)$ is real valued, $\uhat_{-k} = \uhat_{k}^*$, and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k(t) \, e^{i k x} \, ,
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Inserting the projected variable, $\uL$, into the KdV equation and applying the projection operator, $\Proj$, again where necessary produces the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad C_2 = \frac{2 \pi^2 \delup}{3 \lamfac^2}
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. Since all wavenumbers larger than $\Lambda$ have been removed, \eqref{TKdV} represents a {\em finite} dimensional dynamical system, of dimension $\Lambda$ over $\mathbb{C}$. The constants $C_3$ and $C_2$ are the same as before and have been repeated here for convenience.

@@

	More specifically, we assume continuity of the surface displacement, $\eta$, across the ADC \cite{whitham2011linear, rey1992propagation}. Since the propagation speed $c$ changes with depth, waves crossing the ADC must rapidly adjust in wavelength in order to match the oscillation frequency just upstream of the ADC \cite{whitham2011linear, rey1992propagation}. The normalized domain $x \in [-\pi,\pi]$ is scaled on the characteristic wavelength and so there is no change in the dimensionless wave-field $u(x,t)$, giving the condition
\begin{equation}
\label{detmatch}
u(x,t) \vert_{t=0^-} = u(x,t) \vert_{t=0^+}, 
\qquad \mbox{\em Deterministic matching condition}
\end{equation}	
We call \eqref{detmatch} the deterministic matching condition to contrast with its statistical counterpart introduced below. This matching condition is employed in the deterministic simulations of TKdV \eqref{TKdV}. We note that in the alternate formulation mentioned earlier, it is the product $\depth^{1/4} \eta$ that would match at the ADC \cite{johnson1997modern}.

@@

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS[\uL]
\end{align}	
where we have represented the right-hand side by the operator $\RHS[\uL]$. We employ a pseudo-spectral discretization of \eqref{TKdV2}, with de-aliasing applied to the quadratic nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3\Lambda/2$, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$.
For simplicity, we denote these Fourier coefficients $\usqdah_k$,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}
With this discretization, \eqref{TKdV2} can be recast in spectral space as a nonlinear ODE system
\begin{equation}
\label{TKdVDisc}
\td{}{t} \uhat_k =  -\frac{1}{2} C_3 \drat^{-3/2} \, ik \usqdah_k + C_2 \drat^{1/2} \, ik^{3} \uhat_k = \RHSh_k
\end{equation}
The quadratic nonlinearity represented by $\usqdah_k$ mixes the modes during evolution. We note the third-order linear term may become stiff for large $\Lambda$.

@@

With the experimental setup described and the theory outlined, we now present results comparing the two. Unless stated otherwise, all parameters used in the theory are taken directly from their experimental values listed in Table \ref{paramtable}.

@@

	Next, we aim to make the same comparison for free-surface slopes. We note that the derivative, $\partial \eta / \partial x$, is already a dimensionless quantity with a simple physical interpretation, namely the slope of the free surface, whereas the interpretation of $\partial u/\partial \tilde{x}$ is tied to the characteristic values $\ampscale$ and $\lengthscale$.  
We therefore convert the theoretical calculated slopes values back to physical slopes via
\begin{equation}
\pd{\eta}{x} = 2 \pi^{3/2} \frac{\etastd}{\lamupdn} \pd{\uupdn}{\tilde{x}}
\end{equation}
So that the peak experimental wavelength $\lam$ (corresponding to $\freqp = 2$ Hz) would be mapped to the dominating TKdV spectral mode, $k = 1$,  this conversion formula has not been corrected with the extra factor of $\lamfac$ present in \eqref{scales}.

@@

@@

Inserting \eqref{H3convert} and \eqref{H2convert} into \eqref{H3H2ratio}, and using the definitions of $C_2$ and $C_3$ from \eqref{C3C2}, gives the explicit formula
\begin{equation}
\label{SkewPred}
\frac{\skwdn(\eta)} {\vardn(\eta_x) - \varup(\eta_x)} = \frac{1}{3} \epsup^{-3} \dratdn^3 
\end{equation}
This formula links the skewness of the outgoing wave field to the change in the variance of surface slope. In particular, it predicts their ratio to scale as the {\em inverse cube of wave amplitude}, $\epsup^{-3}$, and the {\em cube of the depth ratio}, $\dratdn^3$. It is a rather unexpected relationship, as it would have been difficult to anticipate that the displacement skewness and slope variance, among all possible variable combinations, are so intimately related.
